---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ddir

<!-- badges: start -->
<!-- badges: end -->

This package provides functions for the calculation and documentation of the
potential of drugs to act as perpetrators of clinically relevant drug-drug
interaction (DDI).

## Installation

You can install the development version of ddir like so:

``` r
devtools::install_github("rstrotmann/ddir")
```

## Example

The package provides sample data for a fictional drug, `examplinib`. The
following code shows the relevant drug properties for examplinib, and calculates
the perpetrator DDI risk for direct CYP inhibition (basic method). Both are
provided as markdown-formatted output tables:

```{r example}
library(ddir)

p <- examplinib_parent
property_table(p)
basic_cyp_inhibition_risk_table(p, examplinib_cyp_inhibition_data)
```

See also: `kinetic_cyp_induction_risk_table()`, `mech_stat_cyp_risk_table()`,
`basic_ugt_inhibition_risk_table()`, `transporter_inhibition_risk_table()`

Full documentation can be found together with the source code on
[github](https://github.com/rstrotmann/ddir).
