---
title: "DDI assessment for examplinib"
author: "Author"
date: "Date"
format: pdf
editor: source
---

```{r echo=F, warning=F, info=F, message=F}
library(tidyverse)
library(knitr)
library(ddir)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=TRUE, fig.width=4, fig.height=3, fig.align="center")
```

```{r ompound_data}
perp <- perpetrator(
  "examplinib",
  dose = 50, imaxss = 566, mw = 420.5, oral = TRUE, fu = 0.046, fa = 1,
  fg = 1, rb = 0.9, ka = 0.1,
  source = c(dose = "clinical dose", imaxss = "study 001", fu = "study 002",
             rb = "study 003")
)

metab <- perpetrator(
  "M1",
  dose = 50, imaxss = 566, mw = 420.5, oral = TRUE, fu = 0.046, fa = 1,
  fg = 1, rb = 0.9, ka = 0.1,
  source = c(dose = "clinical dose", imaxss = "study 001", fu = "study 002",
             rb = "study 003")  
)
```

```{r}
cyp_inh <- inhibitor(
  data = tribble(
    ~target, ~ki, ~source,
    "CYP2C9",  7.01,  "study 001",
    "CYP2C19", 48.7, "study 001",
    "CYP3A4",  100, "study 001"
  ),
  object = "examplinib"
)

ugt_inh <- inhibitor(
  data = tibble::tribble(
      ~target,  ~ki,     ~source,
     "UGT1A1",   15, "study 009",
     "UGT1A3",   15, "study 009",
     "UGT1A4",   15, "study 009",
     "UGT1A6",   15, "study 009",
     "UGT1A9",  3.8, "study 009",
     "UGT2B7",   15, "study 009",
    "UGT2B15",   15, "study 009",
    "UGT2B17",  6.1, "study 009",
  ),
  object = "examplinib"
)

transporter_inh <- inhibitor(
  data = tibble::tribble(
    ~target, ~ic50,     ~source,
      "Pgp",  0.41, "study 005",
     "BCRP",   1.9, "study 005",
     "OCT1",   2.3, "study 006",
  "OATP1B1",   177, "study 006",
  "OATP1B3",    35, "study 006",
     "OAT1",   271,          NA,
     "OAT3",   300,          NA,
     "BSEP",  12.8,          NA,
     "OCT2",    67, "study 006",
    "MATE1",   3.6, "study 006",
   "MATE2k",   1.1, "study 006"
  ),
  object = "examplinib"
)

cyp_tdi <- inhibitor(
  data = tibble::tribble(
    ~target, ~ki, ~kinact, ~source,
    "CYP3A4", 0.17,   0.04, "study 001"
  ),
  object = "examplinib"
)

cyp_ind <- inducer(
  data =tibble::tribble(
       ~target, ~emax, ~ec50, ~max_c,     ~source,
      "CYP1A2",     1,    NA,      5, "study 007",
      "CYP2B6",     1,    NA,      5, "study 007",
      "CYP2C8",    NA,    NA,     NA,          NA,
      "CYP2C9",    NA,    NA,     NA,          NA,
     "CYP2C19",    NA,    NA,     NA,          NA,
      "CYP2D6",    NA,    NA,     NA,          NA,
      "CYP3A4",  7.35,  1.64,      3, "study 007"
     ),
  object = "examplinib"
)
```

# Introduction

This drug-drug interaction (DDI) precipitant risks assessment report is based on
the [ICH M12 guidance
document](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf).

All calculations were done using the
[`ddir`](https://github.com/rstrotmann/ddir) package, version `r packageVersion("ddir")`.

# Drug properties

The following physico-chemical, ADME and clinical exposure data were used for
the calculation of the relevant precipitant drug concentrations:

```{r perpetrator_properties, results="asis"}
print(perp)
```

# Key perpetrator concentrations

The following perpetrator concentrations were calculated as described in 
Appendix 1:

```{r results="asis"}
key_conc_table(perp)
```

# DDI risk as inhibitor or inducer of drug-metabolizing enzymes

## Basic modeling of CYP inhibition

### Reversible inhibition

Following the basic modeling approach (refer to Section 2.1.2.1 of the [ICH M12
guideline](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf)),
the relevant metric for the assessment of the direct CYP inhibition risk is
$R=[I]/K_{i,u}$ with the inhibitor concentration $[I]$ being $C_{max,ss,u}$
for hepatic CYP enzymes and $I_{gut}$ for intestinal CYP enzymes.

R values of $R \ge 0.02$ and $R \ge 10$ for hepatic and intestinal enzymes are
considered to indicate a potential clinical risk.

```{r results="asis"}
print(basic_cyp_inhibition_risk(perp, cyp_inh))
```

### Time-dependent inhibition

The risk for time-dependent inhibition (TDI) of CYP enzymes is assessed based on
the formula given in Section '[Time-dependent CYP inhibition]' in Appendix 1,
where $R \ge 1.25$ suggest a clinically relevant DDI potential that requires
further investigation (refer to Section 2.1.2.2 of the [ICH M12
guideline](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf)).

```{r results="asis"}
print(basic_cyp_tdi_risk(perp, cyp_tdi))
```

## Modeling of CYP induction

### Basic mRNA 'fold-change' method

The basic mRNA 'fold-change' approach evaluates whether the maximal change in
CYP mRNA expression is $\ge$ 2-fold at concentrations up to 50-fold above the
expected unbound systemic concentration of the drug (refer to Section 2.1.4.1 of
the [ICH M12 guideline
document](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf)).

Basic modeling as per the ICH M12 guideline results in the following risk
assessment. Data from CYP induction studies that did not cover a concentration
range up to 50-fold over $C_{max,ss,u}$ are flagged in the table. In these
cases, CYP mRNA fold-changes ($E_{max}$) < 2-fold cannot fully exclude a CYP
induction risk.

```{r results="asis"}
print(static_cyp_induction_risk(perp, cyp_ind))
```

### Basic kinetic method

The basic kinetic method for the assessment of CYP induction is based on the
$EC_{50}$ and $E_{max}$ parameters derived from in vitro studies (refer to
Section '[Basic kinetic modeling of CYP induction]' in Appendix 1). For $R < 0.8$,
an in vivo induction risk cannot be excluded:

```{r results="asis"}
print(kinetic_cyp_induction_risk(perp, cyp_ind))
```

## Mechanistic static modeling

Using the mechanistic static modeling approach (refer to Section 7.5.1.2 of the
[ICH M12 guideline](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf), AUC ratios for specific sensitive CYP substrates
are calculated, considering the available in vitro data for both direct and
time-dependent inhibition, and mRNA induction (refer to Section '[Mechanistic static modeling of CYP modulation]' in Appendix 1).
 
Mechanistic static modeling may be used to investigate CYP inhibition alone, or
both inhibition and induction effects. AUC ratios outside the 0.8 to 1.25
interval are considered to indicate a clinical risk.

### CYP inhibition only

```{r results="asis"}
print(mech_stat_cyp_risk(
  perp, cyp_inh, cyp_ind, cyp_tdi, include_induction = FALSE)
)
```

### CYP inhibition and induction

```{r results="asis"}
print(mech_stat_cyp_risk(
  perp, cyp_inh, cyp_ind, cyp_tdi, include_induction = TRUE)
)
```

## Basic modeling of UGT inhibition

The relevant metric for basic modeling of the UGT inhibition risk is
$R=C_{max,ss,u}/K_{i,u}$ (refer to Section 2.1.2.1 of the [ICH M12 guidance document](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf)) for details.

$R>0.02$ is considered to indicate a potential UGT inhibition risk.

Note that in in vitro UGT inhibition studies, often $IC_{50}$ rather than $K_i$
values are reported. Assuming that substrate concentrations close to $K_m$ are
used, $K_i$ is calculated as $K_i = IC_{50}/2$ (refer to Cheng, Prusoff 1973).
 
Basic modeling of UGT inhibition results in the following risk assessment:

```{r results="asis"}
print(basic_ugt_inhibition_risk(perp, ugt_inh))

```

## DDI risk as inhibitor of drug transporters

The metric for the assessment of the drug transporter inhibition risk is
$R=[I]/IC_{50,u}$. For the perpetrator concentrations relevant for the
individual transporters, refer to Section '[Inhibition of drug transporters]' in
Appendix 1.

Note that OCT1 is currently not covered by the M12 guideline.

```{r results="asis"}
print(transporter_inh_risk(perp, transporter_inh))
```

\newpage
# Appendix 1: Calculations and formulae

## Relevant precipitant drug concentrations

### Gut concentration {-}

The maximal gut concentration ($I_{gut}$) for the
orally administered compounds is the administered dose dissolved in 250 ml.

$$I_{gut} = \frac{D} {250}$$

### Systemic concentration {-}

The unbound systemic ($C_{max,ss,u}$) concentration is considered the relevant
precipitant concentration for hepatic enzyme inhibition and induction:

$$C_{max,ss,u}=I_{max,ss} * f_u$$

### Hepatic inlet concentration {-}

The hepatic inlet concentration is considered the relevant perpetrator
concentration for inhibition of the hepatic uptake transporters OATP1B1 and
OATP1B3, and for the hepatic terms in the mechanistic static modeling equation
(refer to Section '[Mechanistic static modeling of CYP inhibition/induction]').

The hepatic inlet concentration is composed of the systemic concentration and
the portal contribution. For orally administered drugs, the portal term is
calculated as:

$$portal\ term = D * \frac{F_a * F_g * k_a}{Q_h * R_B} * 1000\ ng/ml$$

with

* $D$ the administered dose in mg
* $F_a$ the fraction absorbed after oral administration
* $F_g$ the fraction available after gut metabolism
* $k_a$ the absorption rate
* $Q_h$ the hepatic blood flow
* $R_B$ the blood-to-plasma ratio.

The standard hepatic blood flow is assumed as 97 l/h/70 kg or 1.61 l/min/70 kg.

The relevant hepatic inlet ($I_{max,inlet,u}$, also called $I_h$ in the
mechanistic static modeling equations) concentration is the sum of the maximal
systemic plasma concentration and the portal contribution:

$$I_{max,inlet,u}=(C_{max,ss} + portal\ term) * f_u$$

### Enteric concentration {-}

For the parent compound, the villous concentration in the gut ($I_{enteric}$,
also called $I_g$ in the mechanistic static modeling equations) is calculated
as: 

$$I_{enteric,u} = D * \frac{F_a*k_a}{Q_{ent}} *1000\ ng/ml$$

with

* $F_a$ the fraction absorbed after oral administration
* $k_a$ the absorption rate constant
* $Q_{ent}$ the enteric villous blood flow

Note that as per the ICH M12 guideline and [Rostami-Hodjegan and Tucker,
2004](https://doi.org/10.1016/j.ddtec.2004.10.002) the blood-to-plasma
distribution ratio and the plasma binding of the drug are not applicable for the
calculation of the villous concentration.

The standard villous blood flow is assumed as 18 l/h/70 kg or 0.3 l/min/70 kg.

## Basic modeling of enzyme inhibition

### Reversible inhibition

For the basic modeling of direct (reversible) enzyme inhibition, the ratios of
the relevant inhibitor concentration to the $K_{i,u}$ are considered (refer to
Section 2.1.2.1 of the [ICH M12 guidance document](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf)).

For in vitro studies conducted using human liver microsomes, the microsomal
unbound fraction, $f_{u,mic}$ is used to calculate $K_{i,u}$. If unknown, a 
default of 1 is assumed.

$R$ values larger than 0.02 (liver) or 10 (gut), are considered to indicate a
potential clinical enzyme inhibition risk using this method.

#### Liver {-}

$$R=\frac{C_{max,ss,u}}{K_{i,u}}$$

#### Gut wall {-}

$$R_{gut}=\frac{I_{gut}}{K_{i,u}}$$

### Time-dependent CYP inhibition

For the basic modeling of the potential for time-dependent CYP inhibition (TDI),
the following metric is considered:

$$R=\frac {k_{obs} + k_{deg}}{k_{deg}}$$

with

$$k_{obs}=\frac {5 * k_{inact}*C_{max,u}}{K_{I,u} + 5 * C_{max,u}}$$

The CYP degradation constant, $k_{deg}$ is a physiological constant that should
be derived from the scientific literature. In this DDI assessment report,
standard values are used unless otherwise indicated.

Values of $R \ge 1.25$ is considered to indicate a clinically relevant TDI
potential and suggest the need for further investigation.

## Basic mRNA fold-change method method to assess CYP induction

This basic risk assessment evaluates the mRNA induction for a set of hepatocyte
batches from different donors. Increases of CYP enzyme mRNA $\ge$ 2-fold at
at concentrations up to 50-fold above $C_{max,u}$ is considered to indicate a 
clinical risk for CYP induction.

In the context of this assessment only the worst-case donor data is considered.

## Basic kinetic modeling of CYP induction

For the basic kinetic modeling of the CYP induction potential, the following
metric is considered (refer to Section 2.1.4.3 of the ICH M12 guideline):

$$R = \frac {1}{1+d* \frac {E_{max}*10*C_{max,ss,u}}{EC_{50,u} + 10*C_{max,ss,u}}}$$

with $d$ a scaling factor that has a standard value of 1. A different value can
be used if warranted by prior experience with the experimental conditions.

$R \le 0.8$ suggest a relevant in vivo CYP induction potential.

## Mechanistic static modeling of CYP modulation

In this approach, AUC ratios for specific DDI object substrates are projected
based on their known intestinal and hepatic metabolism. Both direct
(competitive) and time-dependent inhibition, as well as enzyme induction are
considered. AUC ratios are calculated according to the below formula (refer to
Section 7.5.1.2 of the ICH M12 guideline):

$$AUCR = \frac{1}{A_g*B_g*C_g*\left(1-F_g\right)+F_g} * \frac{1}{A_h*B_h*C_h*f_m+\left(1-f_m\right)}$$

This calculation is applied for typical probe substrates for which $F_g$, i.e., 
the fraction escaping gut metabolism and $f_m$, i.e., the fraction metabolized 
are known.

Note that the $f_m$ is composed of the overall fraction metabolized for the
respective probe substrate, and the fraction metabolized by the CYP enzyme in
questions:

$$f_m=f_{m,overall} * f_{m,CYP}$$

The individual terms in the AUC calculation are:

### Reversible inhibition {-}

$$A_g = \frac{1}{1+\frac{I_g}{K_i}}$$

$$A_h = \frac{1}{1+\frac{I_h}{K_i}}$$

### Time-dependent inhibition {-}

$$B_g = \frac{k_{deg,g}}{k_{deg,g} + \frac{I_g*k_{inact}}{I_g+K_I}}$$

$$B_h = \frac{k_{deg,h}}{k_{deg,h} + \frac{I_h*k_{inact}}{I_h+K_I}}$$

### Induction {-}

$$C_g = 1 + \frac{d*E_{max}*I_g}{I_g+EC_{50}}$$

$$C_h = 1 + \frac{d*E_{max}*I_h}{I_h+EC_{50}}$$

with the hepatic inlet concentration $I_h=I_{max,inlet,u}$ and the intestinal
concentration $I_g=I_{enteric}$ (see above). $d$ is an induction scaling factor
(assumed to be 1 but can be adjusted based on the experimental conditions).

If the predicted AUC ratio is outside of the 0.8 to 1.25 interval, further
evaluation is required.

## Inhibition of drug transporters

As per the M12 guideline, the metric for the assessment of the drug transporter
inhibition risk is:

$$R=[I]/IC_{50,u}$$

In the respective in vitro studies, the substrate concentration is usually very
low, so that $K_i \approx IC_{50}$ can be assumed. Under common assay conditions,
no protein is added to the medium so that the fraction unbound can be assumed 1,
i.e. $IC_{50} = IC_{50,u}$.

The following relevant precitipant concentrations $[I]$ and regulatory
thresholds of concern apply for the transporters:

| $I$                     | transporter                                                                    | threshold |
| ---                   | ---                                                                            | ---       |
| $I_{gut}$         | P-gp and BRCR when drugs are orally administered                               | 10        |
| $C_{max,ss,u}$    | P-gp and BRCR when drugs are administered parenterally or for drug metabolites | 0.02      |
| $I_{max,inlet,u}$ | hepatic basolateral transporters OCT1, OATP1B1 and OATP1B3                     | 0.1       |
| $C_{max,ss,u}$    | renal basolateral transporters OAT1, OAT3 and OCT2                             | 0.1       |
| $C_{max,ss,u}$    | apical transporters MATE1 and MATE2-K                                          | 0.02      |


Refer to Section '[Relevant precipitant drug concentrations]' for the calculation
of the relevant precipitant concentrations.

# Appendix 2: R Session Info

This document was created using `r sessionInfo()[[1]][["version.string"]]` and
the following packages:

```{r}
si <- sessionInfo()[["otherPkgs"]]

name_vers <- function(item) {
  item <- as.list(item)
  return(c(name=as.character(item[["Package"]]), version=as.character(item[["Version"]])))
}

data.frame(lapply(si, name_vers)) |>
  t() |> 
  kable(row.names = F)
```
