---
title: "DDI assessment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ddi-assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to assess the potential of a drug to act as a
perpetrator of drug-drug interactions with CYP and UGT enzymes and drug
transporters. The live code contained in this document relies on the following R
packages:

```{r setup, message = FALSE, warning = FALSE}
library(ddir)
library(tidyverse)
```

# Regulatory background

In May-2024, ICH published a new harmonized guidance on the assessment of
enzyme- or transporter-mediated drug interactions that will be adopted by the
ICH-abiding regulatory agencies, including FDA, EMA and PMDA.

Currently (Nov-2024), the following guidance documents are provided by the
individual regulatory authorities:

* [M12 Drug Interaction Studies (FDA, Aug-2024)](https://www.fda.gov/media/161199/download)
* [M12 Questions and Answers document (Aug-2024)](https://www.fda.gov/media/180488/download?attachment)
* [ICH M12 Drug Interaction Studies (EMA, May-2024)](https://www.ema.europa.eu/en/documents/scientific-guideline/ich-m12-guideline-drug-interaction-studies-step-5_en.pdf)

# Drug properties

As a general concept, the risk for enzyme or transporter interactions is
evaluated for a given drug exposure level that corresponds to the maximal
unbound clinical exposure in the relevant pharmacokinetic compartment. For
interactions with hepatic enzymes, the unbound maximal plasma concentration at
steady state ($I_{max,ss,u}$) is considered, and for intestinal enzymes as
victim of orally administered drugs, the maximal clinical dose, dissolved in a
volume of 250 mg ($I_{gut}$).

For interactions with basolateral hepatic transporters (OATP1B1, OATP1B3), the
unbound hepatic inlet concentration ($I_{max,inlet,u}$) is considered that is
composed of the $I_{max,ss,u}$ and a portal venous term reflecting the
intestinally absorbed drug escaping gut metabolism.

These concentrations can be automatically derived from the following set of
drug-specific parameters:

| Parameter                                    | Parameter name | Default | Unit  |
| :---                                         | :---:          | :---:   | :---: |
| Molar weight                                 | mw             | NA      | g/mol |
| Clinical dose                                | dose           | NA      | mg    |
| Maximal total plasma concentration ($C_max$) | imaxss         | NA      | ng/ml |
| Fraction unbound ($f_u$)                     | fu             | 1       |       |
| Microsomal unbound fraction ($f_{u,mic}$)    | fumic          | 1       |       |
| Blood-to-plasma concentration ratio ($R_b$)  | rb             | 1       |       |
| Fraction absorbed ($f_a$)                    | fa             | 1       |       |
| Fraction escapting gut metabolism ($f_g$)    | fg             | 1       |       |
| Absorption rate constant ($k_a$)             | ka             | 0.1     | 1/min |
| Solubility                                   | solubility     | Inf     | mg/l  |

The specification of the key compound parameters is therefore the first step in
the formal analysis. In the scope of this package, these key parameters are
aggregated into a 'perpetrator' object. The package contains a sample
perpetrator object for the ficitional drug 'examplinib':

```{r}
examplinib_parent
```

The `ddir` package provides multiple options to create perpetrator objects. In the simplest case, the function `make_perpetrator()` can be used:

```{r}
perp <- make_perpetrator(
  name = "test",
  dose = 100,
  imaxss = 1000,
  mw = 500,
  oral = TRUE,
  fu = 1,
  fumic = 1,
  rb = 1,
  fa = 1,
  fg = 1,
  ka = 0.1,
  solubility = Inf
)
```

Other options are presented in a dedicated vignette (see
`vignette("load-data")`).

To facilitate the generation of R notebook-based reports, `property_table()` can
be used to print a markdown-formatted table of the key compound parameters:

```{r}
property_table(examplinib_parent)
```

# Perpetrator concentrations

The relevant perpetrator concentrations (in $\mu M$) for a perpetrator compound
can be determined using `key_concentrations()` - please see the documentation to
this function for details about the calculations. While this function returns a
named numeric vector of the naked concentrations, a markdown-formatted tabular
output that can be directly embedded in R notebook documents can be generated
using `conc_table()':

```{r}
key_concentrations(examplinib_parent)

conc_table(examplinib_parent)
```

# Direct enzyme inhibition

## CYP enzymes

```{r}
basic_cyp_inhibition_risk(examplinib_parent, examplinib_cyp_inhibition_data)

basic_cyp_inhibition_risk_table(examplinib_parent, examplinib_cyp_inhibition_data)
```

## UGT enzymes

```{r}
basic_ugt_inhibition_risk(examplinib_parent, examplinib_ugt_inhibition_data)

basic_ugt_inhibition_risk_table(examplinib_parent, examplinib_ugt_inhibition_data)
```

# Time-dependent enzyme inhibition

```{r}
basic_cyp_tdi_risk(examplinib_parent, examplinib_cyp_tdi_data)

basic_cyp_tdi_risk_table(examplinib_parent, examplinib_cyp_tdi_data)
```

# CYP induction

## Fold-change method

```{r}
static_cyp_induction_risk(examplinib_parent, examplinib_cyp_induction_data)

static_cyp_induction_risk_table(examplinib_parent, examplinib_cyp_induction_data)
```

## Basic kinetic model

```{r}
kinetic_cyp_induction_risk(examplinib_parent, examplinib_cyp_induction_data)

kinetic_cyp_induction_risk_table(examplinib_parent, examplinib_cyp_induction_data)
```

# Mechanistic-static assessment of CYP-related DDI

```{r}
mech_stat_cyp_risk(examplinib_parent, examplinib_cyp_inhibition_data,
                   examplinib_cyp_induction_data)

mech_stat_cyp_risk_table(examplinib_parent, examplinib_cyp_inhibition_data,
                         examplinib_cyp_induction_data)
```

# Transporter inhibition

```{r}
transporter_inhibition_risk(examplinib_parent, examplinib_transporter_inhibition_data)

transporter_inhibition_risk_table(examplinib_parent, examplinib_transporter_inhibition_data)
```





