---
title: "Load compound-specific data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{load-data} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes how to provide the compound-specific data for the
calculation of the DDI perpetrator risks.

In the first section, different data loading methods are presented, using the
perpetrator drug property object as an example. The second section describes the
requirements for the CYP-, UGT- or transporter inhibition data.

```{r setup, message = FALSE, warning = FALSE}
library(ddir)
library(tidyverse)
```

# Drug properties

All functions within the `ddir` package that evaluate the various DDI risks
expect as their first argument a `perpetrator` object that contains the compound
properties that are needed to calculate the relevant perpetrator concentrations.

The following parameters are required to create a perpetrator object:

| PARAM | REQUIREMENT | DESCRIPTION |
| --- | --- | --- |
| name | required | the name of the compound
| oral | required | logical value (`TRUE` or `FALSE`) to indicate whether the compound is administered orally |
| mw | required | molar weight in g/mol |
| dose | required for oral compounds | clinical dose in mg |
| imaxss | required | total $C_{max}$ in ng/ml |
| fu | optional | fraction unbound, defaults to 1 |
| fumic | optional | microsomal unbound fraction, defaults to 1 |
| rb | optional | blood-to-plasma concentration ratio, defaults to 1 |
| fa | optional | fraction absorbed, defaults to 1 |
| fg | optional | fraction escaping gut metabolism, defaults to 1 |
| ka | optional | absorption rate constant, defaults to 0.1 /min |
| solubility | optional | aqueous solubility in mg/l, defaults to `Inf` |

## Input as data frame

As the most basic option, the compound property data can be hard-coded into a R
script as a data frame and then converted into a perpetrator object:

```{r}
parent_data_frame <- data.frame(
  param = c("name", "oral", "mw", "dose", "imaxss"),
  value = c("examplinib", "TRUE", "492.6", "450", "3530"),
  source = c("", "", "", "", "")
)

parent <- new_perpetrator(parent_data_frame)
parent
```

> Note that the constructor function for perpetrator objects expects all values
to be of character type.

In the above code, only the strictly required fields were provided, the other
fields were automatically set to their respective default values during the
construction of the perpetrator compound.

## Input as tribble

The `tibble` package that is part of the `tidyverse` meta-package provides a
very convenient function, `tribble()`, to construct data frames in a row-wise
fashion. This renders the code much more readable:

```{r}
parent_tribble <- tribble(
  ~param,   ~value,       ~source,
  "name",   "examplinib", "",
  "oral",   "TRUE",       "",
  "mw",     "492.6",     "",
  "dose",   "450",        "",
  "imaxss", "3530",       "" 
)
parent <- new_perpetrator(parent_tribble)
parent
```

## Read from a csv file

Alternatively, the data can also be provided as comma-separated values (csv) in
an external data file, and read using `read_perpetrators()`. The csv file
'examplinib_compounds_single.csv' is included as part of this package for
demonstration purposes:

```{r}
parent_perpetrator <- read_perpetrators(
  fs::path_package("extdata/examplinib_compounds_single.csv", package = "ddir")
)
parent_perpetrator %>%
  class()
```

In the above code, the
`fs::path_package("extdata/examplinib_compounds_single.csv", package = "ddir")`
construct is needed to resolve the path to the sample file within the package
structure. In regular use, the file name should provided here.

The csv file is expected to have 'long table' format: All line must start with
the compound name, followed by the parameter name, the value and (optionally)
source information.

This is what the content of a typical compound file may look like:

```{r echo = FALSE, comment=NA}
cat(examplinib_compounds_single_string)
```

The fields within a line must be separated by commas. All lines starting with
'#' are considered comments and are not evaluated.

## Read from a data chunk within an R notebook

When the DDI risk calculations are done within an R notebook, which we highly
recommend to keep the data, code and report text in a single Rmd document that
can be knitted into a human-readable report, the compound-specific data can be
conveniently provided in the form of data chunks. This is a neat feature
introduced by the
[`knitrdata`](https://journal.r-project.org/articles/RJ-2023-001/) package.

```{r}
library(knitrdata)
```

The text contained in chunks of the type 'data' is assigned as-is to a character
variable by the name defined by 'output.var' in the chunk header when the
document is rendered:

```{r, include=FALSE}
knitr::knit_hooks$set(wrapper = function(before, options) {
  # the original chunk might be indented
  if (is.null(indent <- options$indent)) indent <- ''
  
  # hide the wrapper=TRUE option
  opts <- gsub(', wrapper=TRUE', '', options$params.src)
  
  if (before) {
    # add the header
    sprintf('\n\n%s````\n```{data %s}\n````\n', indent, opts)
  } else {
    # add the footer
    sprintf('\n\n%s````\n```\n````\n', indent)
  }
})
```

```{data output.var = "compounds_text", wrapper = TRUE}
# PARENT
# compound,  param,    value,     source
examplinib,  oral,     TRUE,
examplinib,  mw,       492.6,
examplinib,  dose,     450,       clinical dose
examplinib,  imaxss,   3530,      study 001
examplinib,  fu,       0.023,     study 002
examplinib,  fumic,    1,         default
examplinib,  rb,       1,         study 003
examplinib,  fa,       0.81,      study 003
examplinib,  fg,       1,         default
examplinib,  ka,       0.00267,   unknown

# METABOLITE
# compound,  param,    value,     source
M1,          oral,     FALSE,
M1,          mw,       506.56,
M1,          dose,     NA,
M1,          imaxss,   1038,      study 001
M1,          fu,       0.012,     study 002
M1,          fumic,    1,         default
M1,          rb,       1,         study 002
M1,          fa,       NA,
M1,          fg,       NA,
M1,          ka,       NA,
```

Similar to the above csv file example, the content of this variable can then be
read into a perpetrator object:

```{r}
compounds <- read_perpetrators(
  textConnection(compounds_text)
)
compounds
```

> Note that since the data is not read from a file, the input string variable is
wrapped in a `textConnection()` construct.

> Also note that the above data file contains data for two different compounds,
'examplinib' and 'M1'. The output of the `read_perpetrators()` function is
therefore a list of perpetrator objects in this case.

You will find that the [sample DDI
report](https://github.com/rstrotmann/ddir/blob/main/R/DDI-report_examplinib.Rmd)
for the fictional drug 'examplinib' that is provided as part of the `ddir`
package uses this approach. For your own DDI assessment, feel free to copy that
Rmd file and merely replace the compound-specific data to obtain a basic DDI
report for your drug.

# Enzyme inhibition and induction data

The following sections describe the format in which the _in vitro_ data for
enzyme or transporter inhibition or induction need to be provided to the
respective `ddir` functions. You can use either of the above methods to load the
data. However, as said before, we suggest the input as data chunks within a R
notebook document.

## Direct CYP inhibition data

The `basic_cyp_inhibition_risk()` and `mech_stat_cyp_risk()` functions (and
their `*_table()` siblings) require the $k_i$ values for direct CYP inhibition.
A data frame with the following fields is expected as the the 'cyp_inh'
parameter to these functions:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| cyp | The CYP enzyme (in upper case) |
| ki | the in vitro $k_i$ in µM |
| source | source information for the parameter |

See the below sample data frame for 'examplinib':

```{r}
examplinib_cyp_inhibition_data
```

Use `read_cyp_inhibitor_data()` to load the data from a csv file or
textConnection.

## UGT inhibition data

The `basic_ugt_inhibition_risk()` function relies on the _in vitro_ data for
direct UGT inhibition. A data frame with the following fields is expected as the
'ugt_inh' parameter to these functions:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| ugt | The UGT enzyme (in upper case) |
| ic50 | the in vitro $k_i$ in µM |
| source | source information for the parameter |

See the below sample data frame for 'examplinib':

```{r}
examplinib_ugt_inhibition_data
```

Use `read_ugt_inhibitor_data()` to load the data from a csv file or
textConnection.

## Time-dependent CYP inhibition (TDI) data

The `basic_cyp_tdi_risk()` and `mech_stat_cyp_risk()` functions (and their
`*_table()` siblings) rely on the _in vitro_ data for TDI. A data frame with the
below fields is expected as the 'cyp_tdi' parameter to these functions:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| cyp | The CYP enzyme (in upper case) |
| ki | the in vitro $k_i$ in µM |
| kinact | the inactivation rate in 1/h |
| source | source information for the parameters |

See the below sample data frame for 'examplinib':

```{r}
examplinib_cyp_tdi_data
```

Use `read_tdi_data()` to load the data from a csv file or
textConnection.

## CYP induction data

The `static_cyp_induction_risk()`, `kinetic_cyp_induction_risk()` and
`mech_stat_cyp_risk()` functions (and their `*_table()` siblings) rely on the
_in vitro_ data for CYP induction. A data frame with the below fields is
expected as the 'cyp_ind' parameter to these functions:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| cyp | The CYP enzyme (in upper case) |
| emax | the in vitro emax, i.e., the maximal change in the mRNA level |
| ec50 | the $EC_{50}$ of the induction in µM |
| maxc | the maximal concentration in µM tested in the in vitro study |
| source | source information for the parameters |

See the below sample data frame for 'examplinib':

```{r}
examplinib_cyp_induction_data
```

Use `read_inducer_data()` to load the data from a csv file or
textConnection.

# Transporter inhibition data

The `transporter_inhibition_risk()` function relies on the _in vitro data_ for
direct transporter inhibition to be provided as a data frame to the
'transporter_inh' parameter.

The following fields are expected:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| transporter | The drug transporter |
| ic50 | the in vitro $IC_{50}$ in µM |
| source | source information for the parameter |

See the below sample data frame for 'examplinib':

```{r}
examplinib_transporter_inhibition_data
```
Use `read_transporter_inhibitor_data()` to load the data from a csv file or
textConnection.




