---
title: "Load compound-specific data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{load-data} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes how the compound-specific data can be provided for the
calculation of the DDI perpetrator risks.

In the first sections, different data input options for the drug properties,
(i.e., the parameters needed to calculate the relevant perpetrator
concentrations) are discussed in detail. In the same way, also the CYP-, UGT- or
transporter inhibition data can be provided using these techniques.

```{r setup, message = FALSE, warning = FALSE}
library(ddir)
library(tidyverse)
```

# Drug properties

All functions of the `ddir` package that evaluate the different DDI risks expect
as their first argument a `perpetrator` object that contains all compound
properties needed to calculate the relevant perpetrator concentrations.

The following parameters are needed:

| PARAM | REQUIREMENT | DESCRIPTION |
| --- | --- | --- |
| name | required | The name of the compound |
| oral | required | Logical value (`TRUE` or `FALSE`) to indicate whether the compound is administered orally |
| mw | required | molar weight in g/mol |
| dose | required for oral compounds | clinical dose in mg |
| imaxss | required | total Cmax in ng/ml |
| fu | optional | fraction unbound, default is 1 |
| fumic | optional | microsomal unbound fraction, dafault is 1 |
| rb | optional | blood-to-plasma concentration ratio, default is 1 |
| fa | optional | fraction absorbed, default is 1 |
| fg | optional | fraction escaping gut metabolism, default is 1 |
| ka | optional | absorption rate constant, default is 0.1 /min |
| solubility | optional | solubility of the compound in mg/l, defaults to `Inf` |

## Input as data frame

As the most basic option, the compound property data can be hard-coded into a R
script as a data frame and then converted into a perpetrator object.

> Note that the constructor function for perpetrator objects expects all values
to be of character type:

```{r}
parent_data_frame <- data.frame(
  param = c("name", "oral", "mw", "dose", "imaxss"),
  value = c("examplinib", "TRUE", "492.6", "450", "3530"),
  source = c("", "", "", "", "")
)

parent <- new_perpetrator(parent_data_frame)
parent
```

In the above code, only the strictly required fields were provided, the other
fields were automatically set to their respective default values during the
construction of the perpetrator compound.

## Input as tribble

The `tibble` package that is included in the `tidyverse` meta-package provides a
very convenient function, `tribble()`, to construct data frames in a row-wise
fashion. This renders the code much more readable:

```{r}
parent_tribble <- tribble(
  ~param,   ~value,       ~source,
  "name",   "examplinib", "",
  "oral",   "TRUE",       "",
  "mw",     "492.6",     "",
  "dose",   "450",        "",
  "imaxss", "3530",       "" 
)
parent <- new_perpetrator(parent_tribble)
parent
```

## Read from a csv file

Alternatively, the data can also be provided as comma-separated values (csv) in
an external data file and read using `read_perpetrators()`. The csv file
'examplinib_compounds_single.csv' is included as part of this package for
demonstration purposes:

```{r eval=FALSE}
parent_perpetrator <- read_perpetrators(
  fs::path_package("extdata/examplinib_compounds_single.csv", package = "ddir")
)
parent_perpetrator %>%
  class()
```

In the above code, the
`fs::path_package("extdata/examplinib_compounds_single.csv", package = "ddir")`
construct is needed to resolve the path to the sample file within the package
structure. In normal use cases, the file name should provided here.

## Read from a data chunk within an R notebook

When programming in an R notebook, which we highly recommend to keep the data,
the code and text blocks together in one document, the compound properties can
also be provided in the form of a data chunk. This is a feature introduced by
the `knitrdata` package:

```{r}
library(knitrdata)
```

Chunks with the type 'data' will be assigned as is to the variable provided by
'output.var':

```{r, include=FALSE}
knitr::knit_hooks$set(wrapper = function(before, options) {
  # the original chunk might be indented
  if (is.null(indent <- options$indent)) indent <- ''
  
  # hide the wrapper=TRUE option
  opts <- gsub(', wrapper=TRUE', '', options$params.src)
  
  if (before) {
    # add the header
    sprintf('\n\n%s````\n```{data %s}\n````\n', indent, opts)
  } else {
    # add the footer
    sprintf('\n\n%s````\n```\n````\n', indent)
  }
})
```


```{data output.var = "compounds_text", wrapper=TRUE}
# PARENT
# compound,  param,    value,     source
examplinib,  oral,     TRUE,
examplinib,  mw,       492.6,
examplinib,  dose,     450,       clinical dose
examplinib,  imaxss,   3530,      study 001
examplinib,  fu,       0.023,     study 002
examplinib,  fumic,    1,         default
examplinib,  rb,       1,         study 003
examplinib,  fa,       0.81,      study 003
examplinib,  fg,       1,         default
examplinib,  ka,       0.00267,   unknown

# METABOLITE 
# compound,  param,    value,     source
M1,          oral,     FALSE,
M1,          mw,       506.56,
M1,          dose,     NA,        
M1,          imaxss,   1038,      study 001
M1,          fu,       0.012,     study 002
M1,          fumic,    1,         default
M1,          rb,       1,         study 002
M1,          fa,       NA,
M1,          fg,       NA,
M1,          ka,       NA,    
```

Like in the above example, this variable can then be used to create perpetrator
objects:

```{r}
compounds <- read_perpetrators(
  textConnection(compounds_text)
)
compounds
```

> Note that since the data is not read from a file, the input string variable is
wrapped in a `textConnection()` construct.

> Also note that the above data chunk contains data for two different compounds,
'examplinib' and 'M1'. The output of the `read_perpetrators()` function is a
list of perpetrator objects in this case.

You will find that the [sample DDI
report](https://github.com/rstrotmann/ddir/blob/main/R/DDI-report_examplinib.Rmd)
provided as part of the `ddir` package uses this approach. Feel free to use this
report file and merely exchange the compound-specific data to create a standard
DDI report.

# Enzyme inhibition and induction data

## Direct CYP inhibition data

The `basic_cyp_inhibition_risk()` and `mech_stat_cyp_risk()` functions (and
their `_table()` siblings) rely on the in vitro data for direct CYP inhibition.

The following fields are expected in the respective data frame:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| cyp | The CYP enzyme (in upper case) |
| ki | the in vitro $k_i$ in µM |
| source | source information for the parameter |

Use `read_cyp_inhibitor_data()` to load the data from a csv file or
textConnection.

## UGT inhibition data

The `basic_ugt_inhibition_risk()` function relies on the in vitro data for
direct UGT inhibition.

The following fields are expected in the respective data frame:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| ugt | The UGT enzyme (in upper case) |
| ic50 | the in vitro $k_i$ in µM |
| source | source information for the parameter |

Use `read_ugt_inhibitor_data()` to load the data from a csv file or
textConnection.

## Time-dependent CYP inhibition (TDI) data

The `basic_cyp_tdi_risk()` and `mech_stat_cyp_risk()` functions (and their
`_table()` siblings) rely on the in vitro data for TDI.

The following fields are expected in the respective data frame:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| cyp | The CYP enzyme (in upper case) |
| ki | the in vitro $k_i$ in µM |
| kinact | the inactivation rate in 1/h |
| source | source information for the parameters |

Use `read_tdi_data()` to load the data from a csv file or
textConnection.

## CYP induction data

The `static_cyp_induction_risk()`, `kinetic_cyp_induction_risk()` and `mech_stat_cyp_risk()` functions (and their
`_table()` siblings) rely on the in vitro data for CYP induction.

The following fields are expected in the respective data frame:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| cyp | The CYP enzyme (in upper case) |
| emax | the in vitro emax, i.e., the maximal change in the mRNA level |
| ec50 | the $EC_{50}$ of the induction in µM |
| maxc | the maximal concentration in µM tested in the in vitro study |
| source | source information for the parameters |

Use `read_inducer_data()` to load the data from a csv file or
textConnection.

# Transporter inhibition data

The `transporter_inhibition_risk()` function relies on the in vitro data for
direct transporter inhibition.

The following fields are expected in the respective data frame:

| PARAM | DESCRIPTION |
| --- | --- |
| name | The name of the compound |
| transporter | The drug transporter |
| ic50 | the in vitro $IC_{50}$ in µM |
| source | source information for the parameter |

Use `read_transporter_inhibitor_data()` to load the data from a csv file or
textConnection.




