---
title: "Load compound-specific data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{load-data} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes how the compound-specific data can be provided for the
calculation of the DDI pepetrator risk.

```{r setup, message = FALSE, warning = FALSE}
library(ddir)
library(tidyverse)
```

# Drug properties

All functions of the `ddir` package that evaluate the different DDI risks expect
as their first argument a `perpetrator` object that contains all compound
properties needed to calculate the relevant perpetrator concentrations.

The following parameters are needed:

| PARAM | REQUIREMENT | DESCRIPTION |
| --- | --- | --- |
| name | required | The name of the compound |
| oral | required | Logical value (`TRUE` or `FALSE`) to indicate whether the compound is administered orally |
| mw | required | molar weight in g/mol |
| dose | required for oral compounds | clinical dose in mg |
| imaxss | required | total Cmax in ng/ml |
| fu | optional | fraction unbound, default is 1 |
| fumic | optional | microsomal unbound fraction, dafault is 1 |
| rb | optional | blood-to-plasma concentration ratio, default is 1 |
| fa | optional | fraction absorbed, default is 1 |
| fg | optional | fraction escaping gut metabolism, default is 1 |
| ka | optional | absorption rate constant, default is 0.1 /min |
| solubility | optional | solubility of the compound in mg/l, defaults to `Inf` |

## Input as data frame

As the most basic option, the compound property data can be hard-coded into a
R script as a data frame and then converted into a perpetrator object. The 
constructor function for perpetrator objects expects all values to be of
character type:

```{r}
parent_data_frame <- data.frame(
  param = c("name", "oral", "mw", "dose", "imaxss"),
  value = c("examplinib", "TRUE", "492.6", "450", "3530"),
  source = c("", "", "", "", "")
)

parent <- new_perpetrator(parent_data_frame)
parent
```

Note that in the above code, only the strictly required fields were provided,
the other fields were automatically set to their respective default values.

## Input as tribble

The `tibble` package that is included in the `tidyverse` meta-package provides a
very convenient function, [tribble()], to construct data frames in a row-wise
fashion:

```{r}
parent_tribble <- tribble(
  ~param,   ~value,       ~source,
  "name",   "examplinib", "",
  "oral",   "TRUE",       "",
  "mw",     "492.6",     "",
  "dose",   "450",        "",
  "imaxss", "3530",       "" 
)
parent <- new_perpetrator(parent_tribble)
parent
```

## Read from a csv file

Alternatively, the data can also be provided as comma-separated values (csv) in 
an external data file and read using `read_perpetrators()`. The csv file
'examplinib_compounds_single.csv' is provided for demonstration purposes as part
of this package:

```{r eval=FALSE}
parent_perpetrator <- read_perpetrators(
  fs::path_package("extdata/examplinib_compounds_single.csv", package = "ddir")
)
parent_perpetrator %>%
  class()
```

## Read from a data chunk within an R notebook

When programming in an R notebook, which we highly recommend to keep the data,
the code and text blocks together in one document, the compound properties can
also be provided in a code chunk. This is a feature introduced by the
`knitrdata` package:

```{r}
library(knitrdata)
```

Chunks with the type 'data' will be assigned as is to the variable provided by
'output.var':

```{data output.var = "compounds_text"}
# PARENT
# compound,  param,    value,     source
examplinib,  oral,     TRUE,
examplinib,  mw,       492.6,
examplinib,  dose,     450,       clinical dose
examplinib,  imaxss,   3530,      study 001
examplinib,  fu,       0.023,     study 002
examplinib,  fumic,    1,         default
examplinib,  rb,       1,         study 003
examplinib,  fa,       0.81,      study 003
examplinib,  fg,       1,         default
examplinib,  ka,       0.00267,   unknown

# METABOLITE 
# compound,  param,    value,     source
M1,          oral,     FALSE,
M1,          mw,       506.56,
M1,          dose,     NA,        
M1,          imaxss,   1038,      study 001
M1,          fu,       0.012,     study 002
M1,          fumic,    1,         default
M1,          rb,       1,         study 002
M1,          fa,       NA,
M1,          fg,       NA,
M1,          ka,       NA,    
```

Like in the above example, this variable can then be used to create perpetrator
objects:

```{r}
compounds <- read_perpetrators(
  textConnection(compounds_text))
compounds
```

Note that since the data is not in a file but in a character variable, a
`textConnection()` construct must be used here.

Also note that the above data chunk contains data for two different compounds,
'examplinib' and 'M1'. The output of the `read_perpetrators()` function is 
therefore a list of perpetrator objects.

You will find that the sample DDI report provided as part of the `ddir` package
uses this concept. Feel free to use this report file and merely exchange the
compound-specific data to create a standard DDI report.







