---
output:
  pdf_document: default
  html_notebook: default
---

```{r echo=F, warning=F, info=F, message=F}
library(tidyverse)
library(knitr)
library(knitrdata)
library(ddir)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3, fig.align="center")
```

<!-- The following sections contain compound-specific data for the DDI calculations. -->
<!-- Please replace the sample data with the relevant information. Note that all  -->
<!-- lines must contain the respective number of fields in a comma-separated way. -->
<!-- The 'source' field is for reference source information, i.e., the relevant -->
<!-- clinical or non-clinical study report. It may remain empty without affecting  -->
<!-- the DDI calculations. -->

```{data output.var="compounds_text"}
# PARENT
# compound,  param,    value,     source
examplinib,  type,     parent,
examplinib,  mw,       492.6,
examplinib,  dose,     450,       clinical dose
examplinib,  imaxss,   3530,      study 001
examplinib,  fu,       0.023,     study 002
examplinib,  fumic,    1,         default
examplinib,  rb,       1,         study 003
examplinib,  fa,       0.81,      study 003
examplinib,  fg,       1,         default
examplinib,  ka,       0.00267,   unknown

# METABOLITE 
# compound,  param,    value,     source
M1,          type,     metabolite,
M1,          mw,       506.56,
M1,          dose,     NA,        
M1,          imaxss,   1038,      study 001
M1,          fu,       0.012,     study 002
M1,          fumic,    1,         default
M1,          rb,       1,         study 002
M1,          fa,       NA,
M1,          fg,       NA,
M1,          ka,       NA,    
```

```{data output.var="cyp_inhibition_text"}
# PARENT
# compound, CYP, ki, source
examplinib, CYP1A2,  NA 
examplinib, CYP2B6,  NA,
examplinib, CYP2C8,  11,   study 001
examplinib, CYP2C9,  13.5, study 001
examplinib, CYP2C19, 15,   study 001
examplinib, CYP2D6,  NA,
examplinib, CYP3A4,  12.5, study 001

# METABOLITE
M1,         CYP2C9,  4.4,  study 002
```

```{data output.var="cyp_induction_text"}
# PARENT
# compound, CYP, Emax, EC50, max c, source
examplinib, CYP1A2,  1,    NA,   5,  study 007
examplinib, CYP2B6,  1,    NA,   5,  study 007
examplinib, CYP2C8,  NA,   NA,   NA,
examplinib, CYP2C9,  NA,   NA,   NA,
examplinib, CYP2C19, NA,   NA,   NA,
examplinib, CYP2D6,  NA,   NA,   NA,
examplinib, CYP3A4,   7.35, 1.64, 3,  study 007

# METABOLITE
# compound, CYP, ki, source
M1, CYP1A2,  1,    NA,   5,  study 007
M1, CYP2B6,  6.98, 1.86, 5,  study 007
M1, CYP2C8,  NA,   NA,   NA, 
M1, CYP2C9,  NA,   NA,   NA, 
M1, CYP2C19, NA,   NA,   NA, 
M1, CYP2D6,  NA,   NA,   NA, 
M1, CYP3A4,  22.7, 1.1,  5,  study 007
```

```{data output.var="ugt_inhibition_text"}
# PARENT
# compound, enzyme, IC50, source
examplinib, UGT1A1, 15, study 009
examplinib, UGT1A3, 15, study 009
examplinib, UGT1A4, 15, study 009
examplinib, UGT1A6, 15, study 009
examplinib, UGT1A9, 3.8, study 009
examplinib, UGT2B7, 15, study 009
examplinib, UGT2B15, 15, study 009
examplinib, UGT2B17, 6.1, study 009

# METABOLITE
# compound, enzyme, IC50, source
M1, UGT1A1, 1.1, study 009
M1, UGT1A3, 5.8, study 009
M1, UGT1A4, 6.2, study 009
M1, UGT1A6, 15, study 009
M1, UGT1A9, 3.6, study 009
M1, UGT2B7, 15, study 009
M1, UGT2B15, 9.6, study 009
```

```{data output.var="transporter_inhibiton_text"}
# PARENT
# name,     transporter, IC50, source
examplinib, Pgp,       0.41,  study 005
examplinib, BCRP,      1.9,  study 005
examplinib, OCT1,      2.3,    study 006
examplinib, OATP1B1,   177,   study 006
examplinib, OATP1B3,   35,   study 006
examplinib, OAT1,      271,
examplinib, OAT3,      300,  
examplinib, BSEP,      12.8,
examplinib, OCT2,      67,    study 006
examplinib, MATE1,     3.6,    study 006
examplinib, MATE2k,    1.1,    study 006
```

<!-- End of the compund-specific data section. -->


```{r load_compound_data}
compounds <- read_perpetrators(textConnection(compounds_text))
cyp_inhibition_data <- read_inhibitor_data(textConnection(cyp_inhibition_text))
cyp_induction_data <- read_inducer_data(textConnection(cyp_induction_text))
ugt_inhibition_data <- read_inhibitor_data(textConnection(ugt_inhibition_text))
transporter_inhibition_data <- read_inhibitor_data(textConnection(transporter_inhibiton_text))
```

---
title: "DDI assessment for `r names_string(compounds)`"
author: "Rainer Strotmann"
date: `r format(Sys.Date(), "%d-%b-%Y")`
---

# Introduction

<!-- This is basic introductory text that should suffice in most cases. Feel free to change or extend but keep concise.
The version information should not be changed as it provides a back link to the script template. -->

This assessment of the perpetrator risks for xxx is based on the relevant FDA
and EMA guidance documents (FDA 2020 and EMA 2012).

This output was generated using version xx (dd-dd-dddd) of the DDI assessment
R script.


# Drug properties

The following physico-chemical, ADME and clinical exposure data were used for the
calculation of the maximal gut and portal vein concentrations, and the unbound
systemic concentrations. 

```{r perpetrator_properties}
p <- compounds[[1]]
m <- compounds[[2]]

property_table(p)
property_table(m)
```

# Key perpetrator concentrations

The relevant perpetrator concentrations are calculated as outlined in the
Appendix:

```{r}
conc_table(p)
conc_table(m)
```

# DDI perpetrator risk assessment

## Basic modeling of CYP inhibition

### Reversible inhibition

As per the FDA guideline (FDA, 2020), the relevant metric for the basic modeling of the direct CYP inhibition risk is $R=1+[I]/K_{i,u}$ with the relevant inhibitor concentration $[I]$ being $I_{max,ss,u}$ for hepatic CYP enzymes and $I_{gut}$ for intestinal CYP enzymes.

Thresholds of $R < 1.02$ and $R < 11$ apply for hepatic and intestinal enzymes, respectively. Note that intestinal CYP inhibition is only evaluated for CYP3A4.

Basic modeling as per the FDA guideline results in the following risk assessment for direct CYP inhibition by xxx.

```{r}
basic_cyp_inhibition_risk_table(p, cyp_inhibition_data)
basic_cyp_inhibition_risk_table(m, cyp_inhibition_data)
```

Note that the ratios used in the EMA guidance correspond to the FDA ratios listed above minus 1.

## Modeling of CYP induction

### Basic/fold-change method

The basic (EMA) or fold-change (FDA) methods evaluate whether the maximal fold-change in mRNA expression is > 2-fold at the expected unbound hepatic concentration of the drug.

For the relevant drug concentrations, the FDA guidance suggests considering $30*I_{max,ss,u}$ while the EMA guidance considers $50*I_{max,ss,u}$ for hepatic and $0.1*I_{gut}$ for intestinal induction. It is expected that the concentrations in the respective in vitro assays cover these concentrations.

Basic modeling as per the FDA guideline results in the following risk assessment:

```{r}
kinetic_cyp_induction_risk_table(p, cyp_induction_data, na.rm=T)
kinetic_cyp_induction_risk_table(m, cyp_induction_data, na.rm=T)
```


## Mechanistic static modeling of CYP modulation

As per the FDA guideline (FDA 2020), the relevant metric for mechanistic static modeling of the CYP inhibition potential is the predicted AUC ratio ($AUCR$) for specific probe substrates. A cut-off of $R<1.25$ applies.
 
In the current implementation of this tool (and listed in the table below), only reversible inhibition and induction (as reflected in the A- and C-terms of the AUCR formula, see Appendix) is included. In general, the FDA guidelines advises to investigate mechanistic static models of inhibition (reversible and mechanism-dependent) and induction separately.

### CYP inhibition only

```{r}
mech_stat_cyp_risk_table(p, cyp_inhibition_data, cyp_induction_data, include_induction=F, na.rm=T)
mech_stat_cyp_risk_table(m, cyp_inhibition_data, cyp_induction_data, include_induction=F, na.rm=T)
```

### CYP inhibition and induction

```{r}
mech_stat_cyp_risk_table(p, cyp_inhibition_data, cyp_induction_data, include_induction=T, na.rm=T)
mech_stat_cyp_risk_table(m, cyp_inhibition_data, cyp_induction_data, include_induction=T, na.rm=T)
```

## Basic modeling of UGT inhibition

The relevant metric for basic modeling of the UGT inhibition risk is $R_1=I_{max,ss,u}/K_{i,u}$. For the clinical risk assessment, a cut-off of $R<1.02$ applies.

In in vitro UGT inhibition studies, ususally $IC_{50}$ rather than $k_i$ values are determined. Assuming that a substrate concentration close to $K_m$ is used, $K_i$ is calculated as $K_i = IC_{50}/2$ (refer to Cheng, Prusoff 1973).
 
Basic modeling as per the FDA guideline (FDA 2020) results in the following risk assessment:

```{r}
basic_ugt_inhibition_risk_table(p, ugt_inhibition_data)
basic_ugt_inhibition_risk_table(m, ugt_inhibition_data)

```

## Inhibition of drug transporters

The relevant metric for the assessment of transporter interactions is $R=[I]/K_i$. In in vitro transporter inhibition studies, $IC_{50}$ values are experimentally determined. Since the transporter substrate concentration is ususally kept very low in relation to $K_m$ in order to minimze passive permeation, $K_i = IC_{50}$ can be assumed.

The relevant perpetrator concentrations $[I]$ are: $I_{gut}$ for intestinal P-gp and BRCR, $I_{max,inlet,u}$ for the hepatic basolateral transporters OCT1, OATP1B1 and OATP1B3, and $I_{max,ss,u}$ for the renal basolateral transporters OAT1, OAT3 and OCT2, as well as the apical transporters outside the intestinal mucosa, i.e., hepatic P-gp and BCRP, and MATE1, MATE2-k.

The FDA and EMA guidelines differ in their threshold definitions. The risk assessments according to both guidelines are presented below:

```{r}
transporter_inhibition_risk_table(p, transporter_inhibition_data)
transporter_inhibition_risk_table(m, transporter_inhibition_data)
```

\newpage
# Appendix 1: Calculations and formulae


#### Gut concentration {-}

As per the FDA guideline, the maximal gut concentration ($I_{gut}$) for the parent compound is to be assumed the administered dose dissolved in 250 ml.

$$I_{gut} = \frac{D} {250}$$

#### Systemic concentration {-}

The unbound systemic ($I_{max,ss,u}$) concentrations of the parent compound and the metabolites that are relevant for the DDI potential are derived from the total maximal plasma concentration and the respective unbound fractions:

$$I_{max,ss,u}=I_{max,ss} * f_u$$

#### Hepatic inlet concentration {-}

For the parent compound, the portal contribution to the hepatic inlet concentration is calculated as:

$$portal\ term = D*\frac{F_a*F_g*k_a}{Q_h*R_B}*1000\ ng/ml$$
With $D$ the administered dose in mg, $F_a$ the fraction absorbed after oral administration, $F_g$ the fraction available after gut metabolism, $k_a$ the absorption rate, $Q_h$ the hepatic blood flow and $R_B$ the blood-to-plasma ratio.

The standard hepatic blood flow is 97 l/h/70 kg or 1.61 l/min/70 kg.

The relevant hepatic inlet ($I_{max,inlet,u}$, also called $I_h$ in the mechanistic static modeling equations) concentration is the sum of the maximal systemic plasma concentration and the portal contribution:

$$I_{max,inlet,u}=(I_{max,ss} + portal\ term) * f_u$$

#### Enteric concentration {-}

For the parent compound, the villous concentration in the gut ($I_{enteric}$, also called $I_g$ in the mechanistic static modeling equations) is calculated as: 

$$I_{enteric,u} = D * \frac{F_a*k_a}{Q_{ent}} *1000\ ng/ml$$
with $F_a$ the fraction absorbed after oral administration, $k_a$ the absorption rate, $Q_{ent}$ the enteric villous blood flow and $R_B$ the blood-to-plasma ratio.

Note that as per the FDA guideline (refer to FDA, 2020, Fig. 7, and Rostami-Hodjegan and Tucker, 2004) the blood-to-plasma ratio and the plasma binding of the drug are ignored.

The standard villous blood flow is 18 l/h/70 kg or 0.3 l/min/70 kg.



## Basic modeling of enzyme inhibition

### Reversible inhibition

For the basic modeling of direct (reversible) enzyme inhibition, the ratios of the relevant inhibitor concentration to the $K_i$ are considered (refer to the FDA guidance, FDA 2020, Fig. 1). A cut-off of 1.02 applies.

#### Liver {-}

$$R_1=1-\frac{I_{max,ss,u}}{K_{i,u}}$$

#### Gut wall {-}

$$R_{1,gut}=1-\frac{I_{gut}}{K_{i,u}}$$

<!-- ### Time-dependent enzyme inhibition -->

## Basic modeling of CYP induction

$$R_3 = \frac {1}{1+d* \frac {E_{max}*10*I_{max,u}}{EC_{50} + 10*I_{max,u}}}$$


## Static mechanistic modeling of CYP inhibition/induction

In this approach, AUC ratios for probe substrates are calculated based on their known intestinal and hepatic metabolism. Both direct (competitive) and time-dependent inhibition are considered. The below formula given by the FDA guideline (refer to FDA 2020, Fig. 7) also includes intestinal and hepatic enzyme induction terms ($C_g$ and $C_h$, respectively). At the same time, the guideline states that both inhibition and induction should be considered separately. For the purpose of modeling used here the induction terms are ignored.

$$AUCR = \frac{1}{A_g*B_g*C_g*\left(1-F_g\right)+F_g} * \frac{1}{A_h*B_h*C_h*f_m+\left(1-f_m\right)}$$
The invidual terms are:

#### Reversible inhibition {-}

$$A_g = \frac{1}{1+\frac{I_g}{K_i}}$$

$$A_h = \frac{1}{1+\frac{I_h}{K_i}}$$

#### Time-dependent inhibition {-}

$$B_g = \frac{k_{deg,g}}{k_{deg,g} + \frac{I_g*k_{inact}}{I_g+K_I}}$$

$$B_h = \frac{k_{deg,h}}{k_{deg,h} + \frac{I_h*k_{inact}}{I_h+K_I}}$$

#### Induction {-}

$$C_g = 1 + \frac{d*E_{max}*I_g}{I_g+EC_{50}}$$

$$C_h = 1 + \frac{d*E_{max}*I_h}{I_h+EC_{50}}$$

with the hepatic inlet concentration $I_h=I_{max,inlet,u}$ and the intestinal concentration $I_g=I_{enteric,u}$ (see above). $d$ is an induction scaling factor (assumed to be 1 but can be adjusted).

\newpage
# Appendix 2: R Session Info

This document was created using `r sessionInfo()[[1]][["version.string"]]` and
the following packages:

```{r}
si <- sessionInfo()[["otherPkgs"]]

name_vers <- function(item) {
  item <- as.list(item)
  return(c(name=as.character(item[["Package"]]), version=as.character(item[["Version"]])))
}

data.frame(lapply(si, name_vers)) %>%
  t() %>% 
  kable(row.names = F)
```

