---
output:
  pdf_document: default
  html_notebook: default
---

```{r echo=F, warning=F, info=F, message=F}
library(tidyverse)
library(knitr)
library(knitrdata)
library(ddir)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=TRUE, fig.width=4, fig.height=3, fig.align="center")
```

<!-- The following sections contain compound-specific data for the DDI calculations. -->
<!-- Please replace the sample data with the relevant information. Note that all  -->
<!-- lines must contain the respective number of fields in a comma-separated way. -->
<!-- The 'source' field is for reference source information, i.e., the relevant -->
<!-- clinical or non-clinical study report. It may remain empty without affecting  -->
<!-- the DDI calculations. -->

```{data output.var="compounds_text"}
# PARENT
# compound,  param,    value,     source
examplinib,  oral,     TRUE,
examplinib,  mw,       492.6,
examplinib,  dose,     450,       clinical dose
examplinib,  imaxss,   3530,      study 001
examplinib,  fu,       0.023,     study 002
examplinib,  fumic,    1,         default
examplinib,  rb,       1,         study 003
examplinib,  fa,       0.81,      study 003
examplinib,  fg,       1,         default
examplinib,  ka,       0.00267,   unknown

# METABOLITE 
# compound,  param,    value,     source
M1,          oral,     FALSE,
M1,          mw,       506.56,
M1,          dose,     NA,        
M1,          imaxss,   1038,      study 001
M1,          fu,       0.012,     study 002
M1,          fumic,    1,         default
M1,          rb,       1,         study 002
M1,          fa,       NA,
M1,          fg,       NA,
M1,          ka,       NA,    
```

```{data output.var="cyp_inhibition_text"}
# PARENT
# compound, CYP, ki, source
examplinib, CYP1A2,  NA 
examplinib, CYP2B6,  NA,
examplinib, CYP2C8,  11,   study 001
examplinib, CYP2C9,  13.5, study 001
examplinib, CYP2C19, 15,   study 001
examplinib, CYP2D6,  NA,
examplinib, CYP3A4,  12.5, study 001

# METABOLITE
M1,         CYP2C9,  4.4,  study 002
```

```{data output.var="cyp_tdi_text"}
# compound, CYP, Ki, Kinact, source
examplinib, CYP3A4, 0.17,   0.04, study 001
```

```{data output.var="cyp_induction_text"}
# PARENT
# compound, CYP, Emax, EC50, max c, source
examplinib, CYP1A2,  1,    NA,   5,  study 007
examplinib, CYP2B6,  1,    NA,   5,  study 007
examplinib, CYP2C8,  NA,   NA,   NA,
examplinib, CYP2C9,  NA,   NA,   NA,
examplinib, CYP2C19, NA,   NA,   NA,
examplinib, CYP2D6,  NA,   NA,   NA,
examplinib, CYP3A4,   7.35, 1.64, 3,  study 007

# METABOLITE
# compound, CYP, ki, source
M1, CYP1A2,  1,    NA,   5,  study 007
M1, CYP2B6,  6.98, 1.86, 5,  study 007
M1, CYP2C8,  NA,   NA,   NA, 
M1, CYP2C9,  NA,   NA,   NA, 
M1, CYP2C19, NA,   NA,   NA, 
M1, CYP2D6,  NA,   NA,   NA, 
M1, CYP3A4,  22.7, 1.1,  5,  study 007
```

```{data output.var="ugt_inhibition_text"}
# PARENT
# compound, enzyme, IC50, source
examplinib, UGT1A1, 15, study 009
examplinib, UGT1A3, 15, study 009
examplinib, UGT1A4, 15, study 009
examplinib, UGT1A6, 15, study 009
examplinib, UGT1A9, 3.8, study 009
examplinib, UGT2B7, 15, study 009
examplinib, UGT2B15, 15, study 009
examplinib, UGT2B17, 6.1, study 009

# METABOLITE
# compound, enzyme, IC50, source
M1, UGT1A1, 1.1, study 009
M1, UGT1A3, 5.8, study 009
M1, UGT1A4, 6.2, study 009
M1, UGT1A6, 15, study 009
M1, UGT1A9, 3.6, study 009
M1, UGT2B7, 15, study 009
M1, UGT2B15, 9.6, study 009
```

```{data output.var="transporter_inhibiton_text"}
# PARENT
# name,     transporter, IC50, source
examplinib, Pgp,       0.41,  study 005
examplinib, BCRP,      1.9,  study 005
examplinib, OCT1,      2.3,    study 006
examplinib, OATP1B1,   177,   study 006
examplinib, OATP1B3,   35,   study 006
examplinib, OAT1,      271,
examplinib, OAT3,      300,  
examplinib, BSEP,      12.8,
examplinib, OCT2,      67,    study 006
examplinib, MATE1,     3.6,    study 006
examplinib, MATE2k,    1.1,    study 006
```

<!-- End of the compund-specific data section. -->


```{r load_compound_data}
compounds <- read_perpetrators(textConnection(compounds_text))
cyp_inhibition_data <- read_cyp_inhibitor_data(textConnection(
  cyp_inhibition_text))
cyp_tdi_data <- read_tdi_data(textConnection(cyp_tdi_text))
cyp_induction_data <- read_inducer_data(textConnection(cyp_induction_text))
ugt_inhibition_data <- read_ugt_inhibitor_data(textConnection(
  ugt_inhibition_text))
transporter_inhibition_data <- read_transporter_inhibitor_data(
  textConnection(transporter_inhibiton_text))
```

---
title: "DDI perpetrator risk assessment for `r compound_names_string(compounds)`"
author: "Author name"
date: `r format(Sys.Date(), "%d-%b-%Y")`
---

# Introduction

<!-- This is basic introductory text that should suffice in most cases. Feel free to change or extend but keep concise.
The version information should not be changed as it provides a back link to the script template. -->

This assessment of the drug-drug interaction (DDI) perpetrator risks for
`r compound_names_string(compounds)` is based on the relevant FDA and EMA
guidance documents (refer to
[FDA 2020](https://www.fda.gov/media/134582/download) and
[EMA 2012](https://www.ema.europa.eu/en/documents/scientific-guideline/guideline-investigation-drug-interactions-revision-1_en.pdf)).

All calculations were done using the
[`ddir`](https://github.com/rstrotmann/ddir) package.

# Drug properties

The following physico-chemical, ADME and clinical exposure data were used for
the calculation of the relevant drug concentrations:

```{r perpetrator_properties, results="asis"}
property_table(compounds)
```

# Key perpetrator concentrations

The following perpetrator concentrations were calculated as described in 
Appendix 1:

```{r results="asis"}
conc_table(compounds)
```

# DDI perpetrator risk assessment

## Basic modeling of CYP inhibition

### Reversible inhibition

Following the basic modeling approach (refer to the
[FDA guideline](https://www.fda.gov/media/134582/download)), the relevant metric
for the assessment of the direct CYP inhibition risk is $R=1+[I]/K_{i,u}$ with
the inhibitor concentration $[I]$ being $I_{max,ss,u}$ for hepatic CYP enzymes
and $I_{gut}$ for intestinal CYP enzymes.

Thresholds of $R < 1.02$ and $R < 11$ apply for hepatic and intestinal enzymes,
respectively. Intestinal CYP inhibition is only evaluated for CYP3A4.

```{r results="asis"}
basic_cyp_inhibition_risk_table(compounds, cyp_inhibition_data)
```

Note that the ratios used in the EMA guidance correspond to the FDA ratios
listed above minus 1.

### Time-dependent inhibition

As per the FDA guideline, the risk for time-dependent inhbition (TDI) of CYP
enzymes is assessed based on $R_2$ (see Appendix 1), where $R_2 \geq 1.25$ suggest
a clinically relevant DDI potential that requires further investigation.

```{r results="asis"}
basic_cyp_tdi_risk_table(compounds, cyp_tdi_data, na.rm=T)
```



## Modeling of CYP induction

### Basic static/fold-change method

The basic static (EMA) or fold-change (FDA) methods evaluate whether the maximal
change in mRNA expression after incubation of hepatocytes with the drug
is > 2-fold at the expected unbound hepatic concentration of the drug.

Regarding the relevant maximal drug concentrations to be tested in vitro
($maxc$), the FDA guidance suggests considering $30*I_{max,ss,u}$ while the EMA
guidance considers $50*I_{max,ss,u}$ for hepatic and $0.1*I_{gut}$ for intestinal
induction. It is expected that the concentrations in the respective in vitro
assays cover these concentrations.

Basic modeling as per the FDA guideline results in the following risk assessment:

```{r results="asis"}
static_cyp_induction_risk_table(compounds, cyp_induction_data, na.rm=T)
```

### Basic kinetic method

In the basic kinetic method for the assessment of CYP induction, the relevant
metric is $R_3$ (see fig. 4 of the
[FDA guidance](https://www.fda.gov/media/134582/download) and Appendix 1). For
$R_3 < 0.8$, a potential in vivo induction risk is assumed that needs further
investigation.

```{r results="asis"}
kinetic_cyp_induction_risk_table(compounds, cyp_induction_data, na.rm=T)
```

## Mechanistic static modeling of CYP modulation

As per the FDA guideline, the relevant metric for mechanistic static
modeling of the CYP inhibition potential is the predicted AUC ratio ($AUCR$) for
specific probe substrates (refer to fig. 7 of the FDA guidance and Appendix 1).
$AUCR \ge 1.25$ are considered relevant and may require clinical DDI assessment.
 
In general, the FDA guidelines advises to investigate mechanistic static models
of inhibition (reversible and mechanism-dependent) and induction separately.

### CYP inhibition only

```{r results="asis"}
mech_stat_cyp_risk_table(compounds, cyp_inhibition_data, cyp_induction_data,
                         cyp_tdi_data, include_induction=F)
```

### CYP inhibition and induction

```{r results="asis"}
mech_stat_cyp_risk_table(compounds, cyp_inhibition_data, cyp_induction_data,
                         cyp_tdi_data, include_induction=T, na.rm=T)
```

## Basic modeling of UGT inhibition

The relevant metric for basic modeling of the UGT inhibition risk is
$R_1=I_{max,ss,u}/K_{i,u}$. For the clinical risk assessment, a cut-off of
$R<1.02$ applies.

In in vitro UGT inhibition studies, ususally $IC_{50}$ rather than $k_i$ values
are determined. Assuming that a substrate concentration close to $K_m$ is used,
$K_i$ is calculated as $K_i = IC_{50}/2$ (refer to Cheng, Prusoff 1973).
 
Basic modeling as per the FDA guideline (FDA 2020) results in the following
risk assessment:

```{r results="asis"}
basic_ugt_inhibition_risk_table(compounds, ugt_inhibition_data)

```

## Inhibition of drug transporters

The relevant metric for the assessment of transporter interactions is
$R=[I]/K_i$. In in vitro transporter inhibition studies, $IC_{50}$ values are
experimentally determined. Since the transporter substrate concentration is
ususally kept very low in relation to $K_m$ in order to minimze passive
permeation, $K_i = IC_{50}$ can be assumed.

The relevant perpetrator concentrations $[I]$ are: $I_{gut}$ for intestinal
P-gp and BRCR, $I_{max,inlet,u}$ for the hepatic basolateral transporters OCT1,
OATP1B1 and OATP1B3, and $I_{max,ss,u}$ for the renal basolateral transporters
OAT1, OAT3 and OCT2, as well as the apical transporters outside the intestinal
mucosa, i.e., hepatic P-gp and BCRP, and MATE1, MATE2-k.

The FDA and EMA guidelines differ in their threshold definitions. The risk
assessments according to both guidelines are presented below:

```{r results="asis"}
transporter_inhibition_risk_table(compounds, transporter_inhibition_data)
```

\newpage
# Appendix 1: Calculations and formulae

## Perpetrator concentrations

### Gut concentration {-}

As per the FDA guideline, the maximal gut concentration ($I_{gut}$) for the
parent compound is to be assumed the administered dose dissolved in 250 ml.

$$I_{gut} = \frac{D} {250}$$

### Systemic concentration {-}

The unbound systemic ($I_{max,ss,u}$) concentrations of the parent compound and
the metabolites that are relevant for the DDI potential are derived from the
total maximal plasma concentration and the respective unbound fractions:

$$I_{max,ss,u}=I_{max,ss} * f_u$$

### Hepatic inlet concentration {-}

For the parent compound, the portal contribution to the hepatic inlet
concentration is calculated as:

$$portal\ term = D*\frac{F_a*F_g*k_a}{Q_h*R_B}*1000\ ng/ml$$
With $D$ the administered dose in mg, $F_a$ the fraction absorbed after oral
administration, $F_g$ the fraction available after gut metabolism, $k_a$ the
absorption rate, $Q_h$ the hepatic blood flow and $R_B$ the blood-to-plasma
ratio.

The standard hepatic blood flow is 97 l/h/70 kg or 1.61 l/min/70 kg.

The relevant hepatic inlet ($I_{max,inlet,u}$, also called $I_h$ in the
mechanistic static modeling equations) concentration is the sum of the maximal
systemic plasma concentration and the portal contribution:

$$I_{max,inlet,u}=(I_{max,ss} + portal\ term) * f_u$$

### Enteric concentration {-}

For the parent compound, the villous concentration in the gut ($I_{enteric}$,
also called $I_g$ in the mechanistic static modeling equations) is calculated
as: 

$$I_{enteric,u} = D * \frac{F_a*k_a}{Q_{ent}} *1000\ ng/ml$$
with $F_a$ the fraction absorbed after oral administration, $k_a$ the absorption
rate, and $Q_{ent}$ the enteric villous blood flow.

Note that as per the FDA and EMA guidelines and [Rostami-Hodjegan and Tucker,
2004](https://doi.org/10.1016/j.ddtec.2004.10.002) the blood-to-plasma ratio and
the plasma binding of the drug are not applicable for the villous concentration.

The standard villous blood flow is 18 l/h/70 kg or 0.3 l/min/70 kg.



## Basic modeling of enzyme inhibition

### Reversible inhibition

For the basic modeling of direct (reversible) enzyme inhibition, the ratios of
the relevant inhibitor concentration to the $K_i$ are considered (refer to the
FDA guidance, FDA 2020, Fig. 1). A cut-off of 1.02 applies.

#### Liver {-}

$$R_1=1+\frac{I_{max,ss,u}}{K_{i,u}}$$

#### Gut wall {-}

$$R_{1,gut}=1+\frac{I_{gut}}{K_{i,u}}$$

<!-- ### Time-dependent enzyme inhibition -->

## Time-dependent CYP inhibition

For the basic modeling of the potential for time-dependent CYP inhibition (TDI),
$R_2$ is considered with:

$$R_2=\frac {k_{obs} + k_{deg}}{k_{deg}}$$

and

$$k_{obs}=\frac {50*k_{inact}*I_{max,u}}{K_{I,u} + 50*I_{max,u}}$$

The CYP degradation constant, $k_{deg}$ is a physiological constant that should
be derived from the scientific literature. In this DDI assessment report,
standard values are used unless otherwise indicated.

Values of $R_2 \ge 1.25$ may indicate a relevant TDI potential and need further 
investigation.

## Basic kinetic modeling of CYP induction

$$R_3 = \frac {1}{1+d* \frac {E_{max}*10*I_{max,u}}{EC_{50} + 10*I_{max,u}}}$$


## Static mechanistic modeling of CYP inhibition/induction

In this approach, AUC ratios for probe substrates are calculated based on their
known intestinal and hepatic metabolism. Both direct (competitive) and
time-dependent inhibition are considered. The below formula given by the FDA
guideline (refer to FDA 2020, Fig. 7) also includes intestinal and hepatic
enzyme induction terms ($C_g$ and $C_h$, respectively). At the same time, the
guideline states that both inhibition and induction should be considered
separately.

$$AUCR = \frac{1}{A_g*B_g*C_g*\left(1-F_g\right)+F_g} * \frac{1}{A_h*B_h*C_h*f_m+\left(1-f_m\right)}$$

This calculation is applied for typical probe substrates for which $F_g$, i.e., 
the fraction escaping gut metabolism and $f_m$, i.e., the fraction metabolized 
are known.

Note that the $f_m$ is composed of the overall fraction metabolized for the
respective probe substrate, and the fraction metabolized by the CYP enzyme in
questions:

$$f_m=f_{m,overall} * f_{m,CYP}$$

The individual terms in the AUC calculation are:

### Reversible inhibition {-}

$$A_g = \frac{1}{1+\frac{I_g}{K_i}}$$

$$A_h = \frac{1}{1+\frac{I_h}{K_i}}$$

### Time-dependent inhibition {-}

$$B_g = \frac{k_{deg,g}}{k_{deg,g} + \frac{I_g*k_{inact}}{I_g+K_I}}$$

$$B_h = \frac{k_{deg,h}}{k_{deg,h} + \frac{I_h*k_{inact}}{I_h+K_I}}$$

### Induction {-}

$$C_g = 1 + \frac{d*E_{max}*I_g}{I_g+EC_{50}}$$

$$C_h = 1 + \frac{d*E_{max}*I_h}{I_h+EC_{50}}$$

with the hepatic inlet concentration $I_h=I_{max,inlet,u}$ and the intestinal
concentration $I_g=I_{enteric}$ (see above). $d$ is an induction scaling factor
(assumed to be 1 but can be adjusted).

\newpage
# Appendix 2: R Session Info

This document was created using `r sessionInfo()[[1]][["version.string"]]` and
the following packages:

```{r}
si <- sessionInfo()[["otherPkgs"]]

name_vers <- function(item) {
  item <- as.list(item)
  return(c(name=as.character(item[["Package"]]), version=as.character(item[["Version"]])))
}

data.frame(lapply(si, name_vers)) %>%
  t() %>% 
  kable(row.names = F)
```

